---
title: "VTPEH 6270 - Checkpoint 4: Data Simulation"
author: "Naiya Patel"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This report uses the edible_plants dataset from the TidyTuesday project (2026-02-03)(https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-02-03/edible_plants.csv), which contains observational data on edible plant species, including cultivation characteristics, environmental requirements, and growth-related variables such as days to harvest. The dataset is used to evaluate whether mean harvest time differs across cultivation types using analysis of variance (ANOVA).

# Objective and Approach

## Specific Objective

> Does mean time to harvest differ across cultivation types. 

This report will simulate plausible differences in harvest duration between cultivation groups to examine how effect size, variability, and sample size influence the power of an ANOVA to detect true differences under realistic biological sampling scenarios.

## Data Description


```{r load_csv, include=FALSE, results='hide'}
setwd("/Users/naiyapatel/Desktop/Biostats")
data_edible_plants <- read.csv("edible_plants (1).csv")

library(dplyr)
library(ggplot2)
library(gridExtra)
library(broom)
library(tidyverse) 

# Clean data: extract numeric days_harvest from ranges like "55-75"
df_clean <- data_edible_plants %>%
  # Extract first number from range (e.g., "55-75" -> 55)
  mutate(days_harvest_numeric = as.numeric(gsub("-.*", "", days_harvest))) %>%
  filter(!is.na(days_harvest_numeric) & !is.na(cultivation)) %>%
  select(common_name, cultivation, days_harvest, days_harvest_numeric)

```

The observed differences in harvest time suggest that cultivation type is a 
meaningful predictor of how long plants take to mature. These differences could reflect inherent biological characteristics of each cultivation group, including growth rate, environmental tolerances, and developmental pathways. 
Understanding these relationships through simulation will help us determine 
how large a sample we need to reliably detect such differences in future studies.

```{r data_desc_table, echo=FALSE}
# Data Description Table
data_description <- data.frame(
  Variable_Name = c("days_harvest_numeric", "cultivation"),
  Variable_Type = c("Continuous (numeric)", "Categorical"),
  R_Class = c("numeric", "character"),
  Description = c(
    "Number of days from planting to harvest (extracted from range)",
    "Type of cultivation method used"
  )
)

knitr::kable(data_description, caption = "Data Description Table")
```

## Data Visualization

```{r data_viz, echo=FALSE, fig.width=10, fig.height=6}
# Summary statistics by cultivation type
summary_stats <- df_clean %>%
  group_by(cultivation) %>%
  summarise(
    n = n(),
    mean = mean(days_harvest_numeric, na.rm = TRUE),
    sd = sd(days_harvest_numeric, na.rm = TRUE),
    min = min(days_harvest_numeric, na.rm = TRUE),
    max = max(days_harvest_numeric, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(mean))

knitr::kable(summary_stats, caption = "Summary Statistics by Cultivation Type", digits = 2)

# Box plot
p1 <- ggplot(df_clean, aes(x = cultivation, y = days_harvest_numeric, fill = cultivation)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Days to Harvest by Cultivation Type",
       x = "Cultivation Type", y = "Days to Harvest") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

# Violin plot
p2 <- ggplot(df_clean, aes(x = cultivation, y = days_harvest_numeric, fill = cultivation)) +
  geom_violin(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3) +
  theme_minimal() +
  labs(title = "Distribution with Points",
       x = "Cultivation Type", y = "Days to Harvest") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Description of Plausible Relationships

Based on visual examination, we observe differences in mean harvest time across cultivation types. The relationship is **categorical/group-based**: different cultivation methods have different mean harvest durations. This follows a one-way ANOVA model:

$$Y_{ij} = μ + α_i + ε_{ij}$$

Where:
- $Y_{ij}$ = days to harvest for observation j in group i
- μ = grand mean
- α_i = effect of cultivation type i
- ε_ij ~ N(0, σ²) = random error

## Parameters of Interest
```{r params_table, echo=FALSE}

parameters <- data.frame(
  Parameter = c("Grand Mean", "Effect Size", "Residual Variance", "Number of Groups"),
  Symbol = c("mu", "delta", "sigma^2", "k"),

  
  Description = c(
    "Overall mean days to harvest across all cultivation types",
    "Difference between highest and lowest group means",
    "Within-group variance (noise/biological variability)",
    "Number of cultivation type groups"
  )
)

knitr::kable(parameters, caption = "Parameters of Interest")
```

Based on our edible_plants dataset, the grand mean harvest time across all 10 cultivation types is approximately 67 days, with individual types ranging from 43 days (Chenopodiaceae) to 81 days (Miscellaneous). The observed effect size—the difference between the fastest and slowest cultivation types—is approximately 37 days. Within-group standard deviations range from 0 to 52 days, with Miscellaneous showing the highest variability (sd = 51.63) and Solanum showing the lowest (sd = 0). These empirical values inform our simulation parameter ranges, ensuring our simulations reflect realistic biological scenarios observed in the data.

# Simulation

## Simulation Basis & 3.2 Simulation Output
```{r sim_functions, include=FALSE}
# Function to simulate ANOVA data
simulate_anova <- function(effect_size, noise, sample_size, n_groups = 3) {
  
  grand_mean <- mean(df_clean$days_harvest_numeric, na.rm = TRUE)
  sigma <- sqrt(noise)
  
  group_effects <- seq(-effect_size/2, effect_size/2, length.out = n_groups)
  
  simulated_data <- data.frame(
    group = rep(1:n_groups, each = sample_size),
    days_harvest = numeric(n_groups * sample_size)
  )
  
  for (i in 1:n_groups) {
    idx <- (i-1) * sample_size + 1:sample_size
    simulated_data$days_harvest[idx] <- grand_mean + 
                                        group_effects[i] + 
                                        rnorm(sample_size, 0, sigma)
  }
  
  simulated_data$group <- factor(simulated_data$group, 
                                 labels = paste0("Type_", 1:n_groups))
  
  return(simulated_data)
}

get_summary_stats <- function(simulated_data) {
  
  summary <- simulated_data %>%
    group_by(group) %>%
    summarise(
      n = n(),
      mean = mean(days_harvest),
      sd = sd(days_harvest),
      se = sd / sqrt(n),
      .groups = 'drop'
    ) %>%
    mutate(effect_size_range = max(mean) - min(mean))
  
  fit <- aov(days_harvest ~ group, data = simulated_data)
  anova_result <- broom::tidy(fit)
  
  f_stat <- anova_result$statistic[1]
  p_value <- anova_result$p.value[1]
  
  return(list(
    summary = summary,
    f_stat = f_stat,
    p_value = p_value
  ))
}

run_simulation <- function(effect_size, noise, sample_size, n_groups = 3, seed = NULL) {
  
  if (!is.null(seed)) set.seed(seed)
  
  sim_data <- simulate_anova(effect_size, noise, sample_size, n_groups)
  results <- get_summary_stats(sim_data)
  
  return(list(
    data = sim_data,
    summary = results$summary,
    f_stat = results$f_stat,
    p_value = results$p_value
  ))
}
```

The simulated example shows a ~17-day difference between Type_1 and Type_3, with within-group standard deviations (6.74-8.00 days) matching the moderate variability seen in our actual data. This demonstrates that our simulation effectively captures realistic biological variation while maintaining clear group separation.

```{r, echo = FALSE}
test_sim <- simulate_anova(effect_size = 20, noise = 50, sample_size = 30)
test_summary <- get_summary_stats(test_sim)

knitr::kable(
  test_summary$summary %>% 
    select(group, n, mean, sd, se) %>%
    mutate(across(where(is.numeric), ~round(., 2))),
  caption = "Example Simulation Output: Summary Statistics",
  col.names = c("Group", "Sample Size", "Mean Days", "Std Dev", "Std Error")
)
```

## Simulation Function
```{r sim_test2, echo=FALSE}
test_result <- run_simulation(effect_size = 20, noise = 50, sample_size = 30, seed = 42)

knitr::kable(
  test_result$summary %>%
    select(group, n, mean, sd, se) %>%
    mutate(across(where(is.numeric), ~round(., 2))),
  caption = "Test Simulation 2: Summary Statistics",
  col.names = c("Group", "Sample Size", "Mean Days", "Std Dev", "Std Error")
)
```

## Simulation Automation
```{r sim_automation}
# Define parameter ranges
effect_sizes <- seq(5, 50, length.out = 10)  # 10 effect sizes
noise_levels <- c(25, 50, 100)               # 3 noise levels
sample_sizes <- seq(10, 100, length.out = 10) # 10 sample sizes

# Storage
simulation_results <- list()
results_summary <- data.frame()

# Run all combinations (10 x 3 x 10 = 300 simulations)
set.seed(123)
counter <- 1

for (n_size in sample_sizes) {
  for (eff_size in effect_sizes) {
    for (noise in noise_levels) {
      
      sim <- run_simulation(
        effect_size = eff_size,
        noise = noise,
        sample_size = n_size,
        n_groups = 3
      )
      
      simulation_results[[counter]] <- sim
      
      results_summary <- rbind(results_summary, data.frame(
        simulation_id = counter,
        sample_size = n_size,
        effect_size = eff_size,
        noise = noise,
        f_stat = sim$f_stat,
        p_value = sim$p_value,
        significant = sim$p_value < 0.05,
        mean_diff = sim$summary$effect_size_range[1]
      ))
      
      counter <- counter + 1
    }
  }
}

# Calculate power
results_summary <- results_summary %>%
  group_by(effect_size, noise, sample_size) %>%
  mutate(power = mean(significant)) %>%
  ungroup()

cat("Total simulations completed:", nrow(results_summary), "\n")
```

Across all 300 simulations, we observe a clear pattern: statistical power is highly dependent on the interplay between effect size, sample size, and noise. Under low noise conditions (σ² = 25), even modest effect sizes (15-20 days) achieve 80% power with sample sizes around n = 30-40. However, under high noise conditions (σ² = 100), achieving the same power requires substantially larger effect sizes (25-30 days) or sample sizes (n ≥ 50-60). This demonstrates that the variability observed in our data (particularly in Miscellaneous and Brassica groups) has important implications for study design: reducing measurement error is as critical as increasing sample size when attempting to detect true biological differences between cultivation types.

# Visualization

```{r heatmap_viz, fig.width=12, fig.height=8, echo = FALSE}
# Heatmap by noise level
heatmap_data <- results_summary %>%
  distinct(sample_size, effect_size, noise, power)

p_heatmap <- ggplot(heatmap_data, aes(x = effect_size, y = sample_size, fill = power)) +
  geom_tile(color = "white", linewidth = 0.5) +
  facet_wrap(~noise, labeller = labeller(noise = c("25" = "Low Noise (σ² = 25)",
                                                      "50" = "Medium Noise (σ² = 50)",
                                                      "100" = "High Noise (σ² = 100)"))) +
  scale_fill_viridis_c(name = "Power", limits = c(0, 1)) +
  labs(
    title = "Statistical Power Across Parameter Space",
    subtitle = "One-way ANOVA: Cultivation Type Effect on Harvest Time",
    x = "Effect Size (Days Between Groups)",
    y = "Sample Size per Group"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  )

print(p_heatmap)
```

```{r line_viz, fig.width=12, fig.height=6, echo = FALSE}
# Power curves by sample size
p_lines <- ggplot(heatmap_data, aes(x = effect_size, y = power, 
                                    color = factor(sample_size))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2, alpha = 0.6) +
  facet_wrap(~noise, labeller = labeller(noise = c("25" = "Low Noise (σ² = 25)",
                                                      "50" = "Medium Noise (σ² = 50)",
                                                      "100" = "High Noise (σ² = 100)"))) +
  scale_color_viridis_d(name = "Sample Size") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Power Curves: Effect Size vs. Statistical Power",
    x = "Effect Size (Days Between Groups)",
    y = "Statistical Power",
    color = "Sample Size"
  ) +
  theme_minimal()

print(p_lines)
```

# Interpretation

## Key Findings
```{r interpret}
min_power <- min(heatmap_data$power)
max_power <- max(heatmap_data$power)

# Find smallest effect size for 80% power
power_80 <- heatmap_data %>%
  filter(power >= 0.80) %>%
  arrange(effect_size) %>%
  slice(1)

cat("Minimum power observed:", round(min_power, 3), "\n")
cat("Maximum power observed:", round(max_power, 3), "\n")
if (nrow(power_80) > 0) {
  cat("Smallest effect size to achieve 80% power:", round(power_80$effect_size[1], 2), "days\n")
}
```

From these simulations, we observe:

1. **Effect of Effect Size**: Power increases dramatically with effect size. Small differences (5-10 days) are difficult to detect; larger differences (40-50 days) achieve high power across all conditions.

2. **Effect of Sample Size**: Larger sample sizes substantially increase power. With n = 10-20, only large effects are detectable. With n = 80-100, even small effects achieve good power.

3. **Effect of Noise/Variability**: Higher biological variability reduces power. Under high noise, substantially larger samples are needed to achieve adequate power.

4. **Practical Recommendations**: To reliably detect meaningful differences in harvest time between cultivation types:
   - Minimum sample size: n ≥ 30-40 per group
   - Aim for n ≥ 50 if substantial variability is expected
   - Focus on cultivation types differing by at least 15-20 days for meaningful outcomes

# AI Use Disclosure Statement

This document was created using Claude to assist with generating the R code structure, debugging data handling issues, and creating the simulation functions. The author reviewed and verified all code, selected appropriate parameter ranges, and interpreted all results.
